Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-03-05 main.Rmd
figure a) and f) kept only for showing harmony and clashes

We will discuss the effect of the interaction of cyclic granularities in this section.<!--In Section \ref{sec:synergy}, we discussed how pairs of granularities can have empty combinations either due to structure of calendar, event location or duration or due to the way they are built. In this section, we will see how these empty combinations affect the visualization when a dependent variable is plotted against these granularities.--> For illustration, distribution of half-hourly electricity consumption of Victoria is plotted across different time granularities in each of the panel in
\autoref{fig:allFig}. \autoref{fig:allFig} (a) shows the letter value plot across days of the month faceted by months like January, March and December. \autoref{fig:allFig} (c) shows box plot  across days of the year by the 1st, 15th, 29th and 31st days of the month. \autoref{fig:allFig} (d) showing violin plot across days of the month faceted by week of the month. \autoref{fig:allFig} (e), variations across week of the year conditional on week of the month can be observed through a ridge plot and \autoref{fig:allFig} (f) shows decile plots across  day of the year and month of the year.

Clearly, in \autoref{fig:allFig}, we observe that some choices of cyclic time granularities work and others do not. In \autoref{fig:allFig} (c), there will be no observations for some combinations day-of-month and day-of-year. In particular, the 1st day of the month can never correspond to 2nd, 3rd or 4th day of the year. On the contrary, for \autoref{fig:allFig} (a), we will not have any combinations with zero combinations because every day-of-week can occur in any month-of-year. Thus the graphs that don't work are those where many of the combination sets are empty. In other words, if there are levels of cyclic granularities plotted across x-axis which are not spanned by levels of cyclic granularities plotted across facets or vice versa, we will have empty sets leading to potential ineffective graphs. We hypothesize that the synergy of these cyclic granularities are thus playing a role while deciding if the resulting plot would be a good candidate for exploratory analysis.

We redefine harmony and clashes as follows: harmonies are pairs of granularities that aid exploratory data analysis, and clashes are pairs that are incompatible with each other for exploratory analysis. As a result, we should avoid plotting clashes as they hinder the exploratory process by having missing combinations of time granularities in each panel.

```{r allFig, fig.height=5.5, out.width="50%", fig.align= 'left',echo=FALSE, eval=TRUE, warning=FALSE,message = FALSE, fig.cap="Various probability distribution plots of electricity consumption data of Victoria from 2012 to 2014. (a) Letter value plot by day-of-month and month-of-year (b) Decile plot by hour-of-day and day-of-week (c) Box plot by day-of-year and day-of-month, (d) Violin plot of day-of-month and week-of-month, (e) Ridge plot by week-of-month and week-of-year, (f) Decile plot by day-of-year and month-of-year. Only plots (a) and (b) show harmonised time variables.", fig.show = 'hold'}

@@

par(mfrow = c(3, 2))

@@

VIC %>%
  create_gran("day_week") %>%
  filter(day_week %in% c("Mon", "Tue", "Sat", "Sun")) %>%
  prob_plot("day_week", "hour_day",
    response = "Demand",
    plot_type = "quantile",
    symmetric = FALSE
  ) +
  ylab("Electricity Demand [KWh]") +
  ggtitle("(b) Decile plot")

VIC %>%
  create_gran("day_month") %>%
  filter(day_month %in% c(1, 15, 29, 31)) %>%
  prob_plot("day_month",
    "day_year",
    response = "Demand",
    plot_type = "boxplot"
  ) +
  ylab("Electricity Demand [KWh]") +
  scale_x_discrete(breaks = seq(0, 366, 60)) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 7, margin = margin())
  ) +
  ggtitle("(c) Box plot")

VIC %>%
  create_gran("week_month") %>%
  filter(week_month %in% c(1, 2, 4)) %>%
  prob_plot("week_month",
    "day_month",
    response = "Demand",
    plot_type = "violin", alpha = 0.3
  ) + ylab("") +
  xlab("Days of the Month") +
  theme(legend.position = "bottom", strip.text = element_text(size = 7, margin = margin())) + scale_x_discrete(breaks = seq(0, 31, 5)) +
  scale_y_continuous(breaks = seq(2000, 9000, 2000)) +
  ggtitle("(d) Violin plot")

VIC %>%
  create_gran("week_month") %>%
  create_gran("week_year") %>%
  filter(week_month %in% c(1, 2, 5), week_year %in% c(1:20)) %>%
  prob_plot("week_year",
    "week_month",
    response = "Demand",
    plot_type = "ridge"
  ) +
  xlab("Electricity Demand [KWh]") +
  ylab("Weeks of the Year") +
  scale_x_continuous(breaks = seq(2000, 10000, 3000)) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 7, margin = margin())
  ) +
  ggtitle("(e) Ridge plot")

VIC %>%
  create_gran("month_year") %>%
  filter(month_year %in% c("Jan", "Jul", "Nov")) %>%
  prob_plot("month_year",
    "day_year",
    response = "Demand",
    plot_type = "quantile",
    quantile_prob = seq(0.1, 0.9, 0.1),
    symmetric = FALSE
  ) +
  # scale_x_discrete(breaks = seq(1, 336, 60)) +
  ylab("") +
  xlab("Day of the Year") + theme(
    legend.position = "none",
    strip.text = element_text(
      size = 7,
      margin = margin()
    )
  ) +
  ggtitle("(f) Decile plot")

>>>>

We will discuss the effect of the interaction of cyclic granularities in this section.<!--In Section \ref{sec:synergy}, we discussed how pairs of granularities can have empty combinations either due to structure of calendar, event location or duration or due to the way they are built. In this section, we will see how these empty combinations affect the visualization when a dependent variable is plotted against these granularities.--> For illustration, distribution of half-hourly electricity consumption of Victoria is plotted across different time granularities in each of the panel in
\autoref{fig:allFig}. \autoref{fig:allFig} (a) shows the letter value plot across days of the month faceted by months like January, March and December. <!---\autoref{fig:allFig} (c) shows box plot  across days of the year by the 1st, 15th, 29th and 31st days of the month. \autoref{fig:allFig} (d) showing violin plot across days of the month faceted by week of the month. \autoref{fig:allFig} (e), variations across week of the year conditional on week of the month can be observed through a ridge plot and \autoref{fig:allFig}---> (b) shows decile plots across day of the year and month of the year.

Clearly, in \autoref{fig:allFig}, we observe that choice (a) of cyclic time granularities work but (b) does not work. In (b) we can not compare the distributions across different facets becuase every day-of-year cannot occur for every month. For example, 21st day of the year can never correspond to the month March or December of any year.
<!---In \autoref{fig:allFig} (c), there will be no observations for some combinations day-of-month and day-of-year. In particular, the 1st day of the month can never correspond to 2nd, 3rd or 4th day of the year. On the contrary, for \autoref{fig:allFig}---> In (a) we will not have any combinations because every day-of-week can occur in every month-of-year. <!---Thus the graphs that don't work are those where many of the combination sets are empty.--->

In other words, if there are levels of cyclic granularities plotted across x-axis which are not spanned by levels of cyclic granularities plotted across facets or vice versa, we will have empty sets leading to potential ineffective graphs. We hypothesize that the synergy of these cyclic granularities are thus playing a role while deciding if the resulting plot would be a good candidate for exploratory analysis.

We redefine harmony and clashes as follows: harmonies are pairs of granularities that do not contain emptry combinations and thus aid exploratory data analysis, and clashes are pairs that do not aid the process of exploratory analysis because they contain empty combinations. As a result, we should avoid plotting clashes.<!---as they hinder the exploratory process by having missing combinations of time granularities in each panel.--->

```{r allFig, fig.height=5.5, out.width="50%", fig.align= 'left',echo=FALSE, eval=TRUE, warning=FALSE,message = FALSE, fig.cap="Various probability distribution plots of electricity consumption data of Victoria from 2012 to 2014. (a) Letter value plot by day-of-month and month-of-year (b) Decile plot by day-of-year and month-of-year. Plot (a) and (b) show harmony and clashes respectively.", fig.show = 'hold'}

@@

par(mfrow = c(1, 2))

@@

# VIC %>%
#   create_gran("day_week") %>%
#   filter(day_week %in% c("Mon", "Tue", "Sat", "Sun")) %>%
#   prob_plot("day_week", "hour_day",
#     response = "Demand",
#     plot_type = "quantile",
#     symmetric = FALSE
#   ) +
#   ylab("Electricity Demand [KWh]") +
#   ggtitle("(b) Decile plot")
# 
# VIC %>%
#   create_gran("day_month") %>%
#   filter(day_month %in% c(1, 15, 29, 31)) %>%
#   prob_plot("day_month",
#     "day_year",
#     response = "Demand",
#     plot_type = "boxplot"
#   ) +
#   ylab("Electricity Demand [KWh]") +
#   scale_x_discrete(breaks = seq(0, 366, 60)) +
#   theme(
#     legend.position = "bottom",
#     strip.text = element_text(size = 7, margin = margin())
#   ) +
#   ggtitle("(c) Box plot")
# 
# VIC %>%
#   create_gran("week_month") %>%
#   filter(week_month %in% c(1, 2, 4)) %>%
#   prob_plot("week_month",
#     "day_month",
#     response = "Demand",
#     plot_type = "violin", alpha = 0.3
#   ) + ylab("") +
#   xlab("Days of the Month") +
#   theme(legend.position = "bottom", strip.text = element_text(size = 7, margin = margin())) + scale_x_discrete(breaks = seq(0, 31, 5)) +
#   scale_y_continuous(breaks = seq(2000, 9000, 2000)) +
#   ggtitle("(d) Violin plot")
# 
# VIC %>%
#   create_gran("week_month") %>%
#   create_gran("week_year") %>%
#   filter(week_month %in% c(1, 2, 5), week_year %in% c(1:20)) %>%
#   prob_plot("week_year",
#     "week_month",
#     response = "Demand",
#     plot_type = "ridge"
#   ) +
#   xlab("Electricity Demand [KWh]") +
#   ylab("Weeks of the Year") +
#   scale_x_continuous(breaks = seq(2000, 10000, 3000)) +
#   theme(
#     legend.position = "bottom",
#     strip.text = element_text(size = 7, margin = margin())
#   ) +
#   ggtitle("(e) Ridge plot")

VIC %>%
  create_gran("month_year") %>%
  filter(month_year %in% c("Jan", "Jul", "Nov")) %>%
  prob_plot("month_year",
    "day_year",
    response = "Demand",
    plot_type = "quantile",
    quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9),
    symmetric = FALSE
  ) +
  # scale_x_discrete(breaks = seq(1, 336, 60)) +
  ylab("") +
  xlab("Day of the Year") + theme(
    legend.position = "bottom",
    strip.text = element_text(
      size = 7,
      margin = margin()
    )
  ) +
  ggtitle("(b) Decile plot")
